Ribosome accumulation during early phase resistance training in humans

### 2022-02-03 manuscript.Rmd
last rev

<!-- The formatting of the manuscript is based on faseb journal, find csl-files at https://github.com/citation-style-language/styles -->

@@

<div custom-style="correspondence_style">
&ast; Correspondance

@@

**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. <span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from</span> vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline and 48 hrs after sessions 1, 4, 5, 8, 9, 12</span> and after eight days of de-training and from a non-training control group (n = 8). In addition, muscle thickness was measured before and after training and control periods.

**Results:** Training led to muscle growth (`r hypertrophy` vs. control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions. Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline). After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session). After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.

@@

For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no robust differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For rpS6, de-training did not affect protein levels, which remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

@@

In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1). In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).

@@

# Discussion
Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels to play a role in regulation of rRNA transcription regulation in response to RT.
Total RNA increases <span custom-style="CHANGE">were not affected</span> by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
and muscle hypertrophy
[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
@RN1755; <!-- Stec Ribosome bio higher in high responders -->
@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->

Total RNA seems to be a valid proxy marker of ribosomal density, as most of the RNA is assumed to be ribosomal RNA
[@RN2223],<!-- Young  -->
which in turn is a valid marker of translational capacity
[@RN2145].
Several studies have shown that total RNA content is altered by RT
<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
@RN1897<!-- 11% 36 sessions young -->],
as was also the case in the present data set.
However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four <span custom-style="CHANGE">sessions</span>, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
[@RN1809;
@RN2358], and may be essential for preparing muscle fibers for subsequent growth
[@RN1809;<!-- Correlation between increase and muscle growth -->
@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
After the 8^th^ session, no meaningful increase or decrease were observed for total RNA or rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes. Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels. This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration --> <span custom-style="CHANGE">This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell. As such, indirect measures of translational capacity through the density of ribosomes in periods of marked muscle hypertrophy may mask an absolute increase in ribosomes.</span>

The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
Individuals with higher rates of accumulation showed larger accretion of muscle mass.
This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass 
[@RN1644;<!-- Figueriedo  -->
@RN1897;
@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
<!-- Reidy correlation between total RNA and muscle hypertrophy-->
differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
[@RN1755]<!-- Stec 2016  -->
an contribute to explain RT volume-dependent changes in muscle mass and strength
[@RN2358].
In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1755]<!-- Stec specific pol I inhibition -->
but not all studies
[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
Whereas aged muscle display higher levels of total RNA at rest 
[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
they show reduced changes in total RNA levels in response to RT
[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized, newly synthesized ribosomes in muscle hypertrophy although we acknowledge that the present study is limited in terms of probing any such heterogeneity.</span>
Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.

<!-- rpS6 dissconect from total RNA levels. -->
In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6. Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period. Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for. Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT. <span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam2007]. This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam2007]. Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined. However, a disconnect between RPS6 and total RNA after de-training could be explained by a greater number of myonuclei supporting a higher constitutive expression of ribosomal proteins. Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the myofibers ribosomal machinery [@murach2022]. However, the present study's observations are limited to a single ribosomal protein and lack data supporting RT-induced myonuclear accretion. A parallel mechanism that could help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis. 
[@RN2588].<!-- Warner 2009 General description of   -->  
</span>

UBF levels robustly explained total RNA levels over the entire course of the intervention. These analyses were done with the number of sessions accounted for, allowing unbiased estimates. Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
From a mechanistic perspective, UBF is an important transcription factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA Pol I
[@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
[@RN1632; <!-- Nader rapamycin block UBF availability  -->
@RN2564].<!-- Hannan 2003 -->
<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1. The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
[@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
Interestingly the availability of UBF *per se* has been shown to regulate rRNA transcription
[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis. <span custom-style="CHANGE"> However, the lack of measurements of UBF in the context of active chromatin interaction may have inhibited us from further explaining the role of UBF in response to the de-training period.</span>

After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
<span custom-style="CHANGE">The magnitude of the decrease in total RNA ($\sim$ 20\%) was similar as observed in untrained muscle in response to complete unloading during a similar time frame (7-10 days) [@gamrin1998; @figueiredo2020]. This may indicate that the increase in total RNA induced by RT is more easily lost than total RNA levels measured during homeostatic conditions (removal of 2-3 sessions of RT during eight days vs. limb immobilization). In addition to reduced synthesis of new ribosomes, degradation of existing ribosomes (ribophagy) likely contributes to the de-training effect seen on total and rRNA levels. Such degradation is highly sensitive to inactivity evident from rats subjected to hindlimb suspension [@figueiredo2020]. The relative contribution to the number of functional ribosomes of pathways regulating synthesis and degradation of ribosomes remains to be determined. Future studies will likely show that the balance between ribosome biogenesis and ribophagy is specific to cell types and stimuli. Such complexity likely exists as both selective and non-selective pathways for ribophagy exists, activated by different upstream mechanisms [@beese2020] in addition to different effectors controlling ribosome biogenesis. For example, in skeletal muscle and in contrast to inactivity-induced RNA degradation [@figueiredo2020], Kim et al. reported that impaired translational capacity in a model of cancer cachexia was primarily driven by a reduced rDNA transcription [@kim2021].</span>

>>>>

@@

<div custom-style="correspondence_style">
&ast; Correspondence

@@

**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. <span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9 and 12</span> and after eight days of de-training, as well as</span> from a non-training control group (n = 8). In addition, muscle thickness was measured before and after training and control periods.

**Results:** Training led to muscle growth (`r hypertrophy` vs. control) with concomitant increases in total RNA, rRNA, UBF, and RPS6 with no differences between volume conditions. Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline). After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session). After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.

@@

For both ribosomal protein S6 (RPS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with RPS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and RPS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For RPS6, de-training did not affect protein levels, which remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while RPS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

@@

In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1). In contrast, no evidence was found for a relationship between total RNA and RPS6 protein levels (Table 1).

@@

# Discussion
Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and RPS6 protein in previously untrained individuals compared to a non-training control group.
These markers accumulated progressively during the initial part of the intervention before they levelled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels to play a role in regulation of rRNA transcription regulation in response to RT.
Total RNA increases <span custom-style="CHANGE">were not affected</span> by weekly fluctuations in training volume. However, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
and muscle hypertrophy
[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
@RN1755; <!-- Stec Ribosome bio higher in high responders -->
@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->

Total RNA seems to be a valid proxy marker of ribosomal density, as most of the RNA is assumed to be ribosomal RNA
[@RN2223],<!-- Young  -->
which in turn is a valid marker of translational capacity
[@RN2145].
Several studies have shown that total RNA content is altered by RT
<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
@RN1897<!-- 11% 36 sessions young -->],
as was also the case in the present data set.
However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four <span custom-style="CHANGE">sessions</span>, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
[@RN1809;
@RN2358], and may be essential for preparing muscle fibres for subsequent growth
[@RN1809;<!-- Correlation between increase and muscle growth -->
@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
After the 8^th^ session, no meaningful increase or decrease were observed for total RNA or rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes. Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated *per* weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels. This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration --> <span custom-style="CHANGE">This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell. As such, indirect measures of translational capacity such as the concentration of total RNA may mask the absolute increase in ribosomes that occurs during periods of muscle hypertrophy.</span>

The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
Individuals with higher rates of accumulation showed larger accretion of muscle mass.
This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass 
[@RN1644;@RN1897;@RN2055],<!-- Figueriedo  --><!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
<!-- Reidy correlation between total RNA and muscle hypertrophy-->
differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
[@RN1755]<!-- Stec 2016  -->
an contribute to explain RT volume-dependent changes in muscle mass and strength
[@RN2358].
In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
[@RN1632;@RN1754;@RN1755]<!-- Nader 2005 blockage of mTOR halts ribo bio and growth --><!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS--><!-- Stec specific pol I inhibition -->
but not all studies
[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA were instead associated with a tendency towards lowered muscle growth.
Overall, the rate of increases in ribosomal density hence seems to be a better predictor for individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
Whereas aged muscle display higher levels of total RNA at rest 
[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
they show reduced changes in total RNA levels in response to RT
[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
Whether these cellular characteristics are related to e.g. differences in fibre type distributions [@RN1707] remains to be determined.
<span custom-style="CHANGE">Furthermore, ribosomal accumulation is unlikely to be the only ribosome-derived trait that is important for training responsiveness. Evidence suggests that mechanical loading may lead to changes in ribosome characteristics [@chaillou2015] potentially leading to heterogeneous tissue-specific ribosome populations [@chaillou2019;  @gupta2014]. Our results support a model where specialized, newly synthesized ribosomes contribute to muscle hypertrophy as the increase in ribosomal content but not absolute levels predicted muscle growth. We, however, acknowledge that the present study does not provide substantial insight into this perspective.</span>
Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.

<!-- RPS6 dissconect from total RNA levels. -->
In the present study, training induced increases in rRNA and total RNA coincided with increases in RPS6. Conversely, changes in total RNA levels and RPS6 in response to de-training did not correspond, as RPS6 protein levels remained elevated after the de-training period. The training induced increases in RPS6 seen in the present study are in agreement to those previously reported in young men
[@RN1642],<!-- Brook 2015 RPS6 increase in resposne to RT over control -->
but not in elderly men and women, where a decrease was observed in response to training despite increases in total RNA and rRNA
[@RN1755].<!-- Stec 2016 RPS6 decreased 22% after four weeks of training in older  -->
Although increases were seen for both RPS6 and and total RNA, RPS6 did not explain variation in total RNA (after controlling for the number of sessions). Together with a disconnect after the de-training period, this suggest that regulation of RPS6 expression and ribosomal RNA transcription display differential temporal responses to RT. <span custom-style="CHANGE">Such difference in temporal regulation of ribosomal RNA and proteins has previously been deduced in cell culture experiments. Briefly, inhibition of protein degradation led to accumulation of ribosomal proteins suggesting that excess amounts of ribosomal proteins are synthesized, imported into the nucleus and rapidly degraded if not incorporated into ribosomes [@lam2007]. This inherent capacity of cells to provide sufficient access to ribosomal proteins also suggests rRNA transcription (and not synthesis of ribosomal proteins) is not rate-limiting during ribosomal biogenesis [@lam2007]. Currently, it remains unknown if RT leads to expansion of the pool of unbound ribosomal proteins in humans. Indeed, in the present study, the disconnect between RPS6 and total RNA after de-training may have been associated with increased numbers of myonuclei, rather than accumulation of RPS6 in existing myonuclei, providing an alternative explanation to the elevated RPS6 levels (which accordingly may have been distributed over a larger number of nuclei). Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the ribosomal pool in myofibers [@murach2022]. Importantly, the present study was not designed to investigate these perspectives, and observations are limited to one single ribosomal protein, with no investigation of RT-induced myonuclear accretion. A parallel mechanism that could futher help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis 
[@RN2588].<!-- Warner 2009 General description of   -->  
</span>

UBF levels robustly explained total RNA levels over the course of the intervention. As these analyses were done while accunting for the number of training sessions, estimates are likely to be unbiased. Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
From a mechanistic perspective, UBF is an important transcription factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA Pol I
[@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
[@RN1632; <!-- Nader rapamycin block UBF availability  -->
@RN2564].<!-- Hannan 2003 -->
<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1. The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
[@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
Interestingly the availability of UBF has been shown to regulate rRNA transcription
[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis. <span custom-style="CHANGE"> However, the lack of measurements of UBF in the context of active chromatin interaction may have inhibited us from further explaining the role of UBF in response to the de-training period.</span>

After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
<span custom-style="CHANGE">The magnitude of the detraining-associated decrease in total RNA ($\sim$ 20\%) is similar to that seen in response to unloading of untrained human muscle over a similar time frame (7-10 days) [@gamrin1998; @figueiredo2020]. This suggests that RT-induced increases in ribosomal content is easily lost, likely driven by a combination of reduced synthesis and enhanced degradation (ribophagy). While the relative contribution of these two remains to be determined, inactivity has previously been shown to induce ribophagy in rat muscle (subjected to hindlimb suspension) [@figueiredo2020]. Future studies should investigate the interrelationship between ribosome biogenesis and ribophagy in response to training and detraining. We suspect that their relative importance for ribosomal homeostasis will vary substantially between different physiological perturbations, as they are regulated by a complex interconnection of stimuli and pathways<!--REF?-->. For example, Kim et al. showed that cancer cachexia is associated with reduced rDNA transcription (and hence reduced translational capacity) [@kim2021], contrasting the inactivity-driven reduction in translational capacity which to a larger degree seems to involve ribophagy [@figueiredo2020].</span>
